---
title: "Stats in R for MBE2017"
author: "Luis Malpica Cruz"
date: "May 12, 2017"
output: html_document
---

##Set up workspace
```{r}
#Set up your workspace
rm(list=ls()) #remove debris from 'Environment'

# Set your working directory manually:
getwd()
setwd(dir = "/Users/Dudisimo/Documents/PhD/TAing/Rworkshop_MBE2017") #diferent for each machine
# or by cliking 'Session' -> 'Set working directory'

# Install (if required) and load packages needed
install.packages("scales")
install.packages("MuMIn")
#After installing them, load them in each R session using...
library("scales")
library("MuMIn")
```

Remember the data analysis cycle: Tidy data -> Transform data -> Visualize -> Model (repeat)
We first load our data into R and verify tidyness

##Load data
```{r, loading data}
barnacles <-read.csv("data/BarnacleData.csv", header=T, sep=",")

#Check structure
str(barnacles)
head(barnacles)
tail(barnacles)
names(barnacles)

#Some easy stats
summary(barnacles)
```

First let's make a simple plot of our barnacle data which has just two continuous variables, strokes/min (dependent variable) and barnacles per quadrat (independent variable).

We will also implement a linear model to analyze the relationship between these variables and make inferences regarding whether their are statistically related.

##Visualization and model implementation: 1st example
```{r, barnacle data}
#The super basic plot, remember that plots date data in the form "independent variable (X), dependent variable (Y)"
plot(barnacles$density, barnacles$strokes) 
#plot(barnacles)

#Let's dress up the plot a bit by adding labels and making our dots look nicer
plot(barnacles$density, barnacles$strokes, xlab = "Barnacles per Quadrat (5 cm2)", ylab = "Feeding Strokes per Minute", main = "Barnacle Density Influences Feeding Frequency", 
     pch = 20, #this changes the symbol type
     cex = 1.5, #this changes the symbol size
     col = alpha("red", 0.5)) #this changes the color and transparency (0.5 = 50% transparent)
                              #helps to see overlapping data points

#Want somemore dressing up? use 'expression' for labels
par(mar=c(5, 4.3, 4, 2) + 0.1) #change printing plot location
plot(barnacles$density, barnacles$strokes, 
     xlab = expression(Barnacle ~ density ~ (5 ~ cm^-2)), 
     ylab = expression(Feeding ~ strokes ~ ( ~ min^-1)),
     main = "Feeding Frequency as a function of barnacle density", pch = 20, #changes symbol type
     cex = 1.5, #for symbol size
     col = alpha("blue", 0.5)) #changes the color and transparency (0.5 = 50% transparent)

text (150, 25, "Barnacles")   #text command adds specified text at a given x, y position (here:150, 25)

#Looks like there's a relationship in our data, let's try making a linear model of the relationship between barnacle density and number of strokes per minute. Remember linear models always take data in the form dependent variable ~ independent variable.

barnaclemodel <- lm(strokes ~ density, barnacles) #the basic model in the form of y = a + bx
plot.new() #now let's look at some diagnostic plots to make sure our model is appropriate
par(mfrow = c(2,2)) #this tells R to set up a 2x2 grid of 4 plots
plot(barnaclemodel) #create the diagnostic plots
dev.off() #turn off the 4x4 grid setup or all your future plots will go onto a grid
summary(barnaclemodel) #the plots look ok, so now let's look at our model summary!

#Summary gives you the estimate for the intercept (a), the slope (b), the standard error for each estimate, and the p-values for those estimates (remember in science the accepted significance threshold is <0.05, i.e. there is an effect of the independent varibale over the dependent variable).
#The R-squared values tells you what proportion of the variation in the data is explained by the independent variable (here barnacle density).s

```

Now let's look at a more complicated data set from our anemone observations which have two continuous variables (reaction time, pool area) and one categorical variable (clone vs. non-clone tentacles).

##Visualization and model implementation: 2nd example
```{r, anemone data}

anemones <-read.csv("data/AnemoneData.csv", header=T, sep=",")
str(anemones)
head(anemones)
names(anemones)
summary(anemones)

#First let's plot our continuous data like before and then make a model
#Always make as many plots as you can
plot(anemones$reaction.time.s ~ anemones$pool.area.cm2)

#Now let's plot continuous versus categorical data, wel'll see it looks different
#R plots continuous vs. categorical data as a box and whisker plot, where the middle line is the mean
plot(anemones$reaction.time.s ~ anemones$different.colony)

#We could also plot this as a bar plot, by first making a vector of the means for each group

#Prep the data
bar.clones<-anemones [ which(anemones$different.colony=="clone"), ]
bar.nonclones<-anemones [ which(anemones$different.colony=="nonclone"), ]
bar.data.clones<- c(bar.clones$reaction.time.s)
bar.data.nonclones<- c(bar.nonclones$reaction.time.s)
bar.means<- c(mean(bar.clones$reaction.time.s), mean(bar.nonclones$reaction.time.s))

#Make the bar plot
anemone.bar<- barplot(bar.means, ylim = c(0,4000), ylab= " Reaction Time (s)", main =  "Reaction Times of Agg. Anemones", col = c("blue","red"))
axis(1, labels=c("Clones", "Nonclones"), at = anemone.bar)
box()
#Add the error bars
# Get standard deviation of each group
# The standard deviations are saved in a matrix of same size 
# as the matrix with midpoints, this is useful for plotting 
# the error bars
stDevs <- matrix(c(sd(bar.data.clones), sd(bar.data.nonclones)), 2)
# Plot the vertical lines of the error bars
# The vertical bars are plotted at the midpoints
segments(anemone.bar, bar.means - stDevs, anemone.bar, bar.means + stDevs, lwd=2)
# Now plot the horizontal bounds for the error bars
# 1. The lower bar
segments(anemone.bar - 0.1, bar.means - stDevs, anemone.bar + 0.1, bar.means - stDevs, lwd=2)
# 2. The upper bar
segments(anemone.bar - 0.1, bar.means + stDevs, anemone.bar + 0.1, bar.means + stDevs, lwd=2)

#---------#--------#---------#--------#
#Basic analysis: T-tests and ANOVAs####

#How could we determine if reaction times in these two groups are different? using a T-test!
#Shows that these groups are significantly different!
t.test(anemones$reaction.time.s ~ anemones$different.colony)

#Now let's make a model including both of these variables
anemonemodel<- lm(anemones$reaction.time.s ~ anemones$different.colony + anemones$pool.area.cm2)
plot.new()
par(mfrow = c(2,2))
plot(anemonemodel)
dev.off()
summary(anemonemodel)

#What if we want to look at more than two groups? Maybe the students collecting the data influenced the results by using different methodology. First let's plot it.
plot(anemones$reaction.time.s ~ anemones$students)

#The data does look like it's different for groups 1 and 4 - why? Because they reported only non-clone data. We can test if more than 2 groups have different means using an ANOVA.

#We need a balanced number of trials for each group for ANOVA, so let's first subset the data to take only groups that did 3 clone and 3 nonclone trials.
an.trim<- (anemones [ which(anemones$students=="Windy&Carly" | anemones$students=="Helen&JessB"| anemones$students=="Gavia&Emma"), ])

anemoneanova <- aov(an.trim$reaction.time.s ~ an.trim$students)
summary(anemoneanova)

#Now let's assume there are small differences in how different groups measured the data that we want to control for, but that we aren't actually interested in. Another example of this is when taking measurements at multiple field sites where conditions vary slightly, and you might want to control for these when getting your  model estimates, but you don't care about the influences of the sites themselves. These nuisance factors are called "random effects" and can be controlled for using a mixed-effects model.

#First we install the mixed model package
install.packages("lme4")
library("lme4")

#Now let's run a mixed model with student group as a random effect and compare with our original model
anemonemodel.2 <- lmer(anemones$reaction.time.s ~ anemones$different.colony + anemones$pool.area.cm2 + (1|anemones$students)) #here +(1|anemone$students) is how you specify a random intercept, that is, if you expect the height of the relationships to vary across sites but not the slope, this is the most basic type of random effect.

#When we compare we see the resulting parameter estimates are only a little different, because we already saw student group doesn't matter, but it can be a lot different!
summary(anemonemodel.2)
summary(anemonemodel)
```

Now we'll look at a totally different data set to study interactions in models:

##Visualization and model implementation: 3rd example (interactions)
```{r, interactions}

# Here we'll look a data on fish age, fish length, and whether it is mature or not
# Variables are age, lengthcm, mature (where 1 = success (mature)), only response variable is categorical
maturedata <-read.csv("C:/Users/Natascia/Desktop/Bamfield/MBE Course/R Workshop/matureWB.csv", header=T, sep=",")
str(maturedata)
head(maturedata)
names(maturedata)
summary(maturedata)
#This time let's attach the data so we don't have to call the whole data set name each time
attach(maturedata)

plot(maturedata)
plot(mature~age)

#Now let's make some models with the variables individually
maturedata.a <- glm(mature~age, family=binomial)
summary(maturedata.a)

maturedata.l <- glm(mature~lengthcm, family=binomial)
summary(maturedata.l)

#But maybe both length and age are important?
maturedata.al <- glm(mature~age+lengthcm, family=binomial)
summary(maturedata.al)

#And we know there is an interaction between age & length, as they grow older, they get longer...
#So we should capture this as an interaction term!
maturedata.alint <- glm(mature~age+lengthcm+age:lengthcm, family=binomial)
summary(maturedata.alint)

#How can we decide which one of these models is best? Using something called the Akaike Information Criterion (AIC). AIC measures the tradeoffs between model fit and support for the variables in the model, with lower values indicating the model that better fits the data. Here we'll use AICc which is adjusted for small sample sizes. You need the package "MuMIn" that we loaded earlier to do this.

AICc(maturedata.a)
AICc(maturedata.l)
AICc(maturedata.al)
AICc(maturedata.alint)

```





