---
title: "Intro to R for MBE2017"
author: "Luis Malpica Cruz"
date: "May 2, 2017"
output: html_document
---

## Showing off R
```{r}

```



## The basics
 This is your working environment in R Studio, here you write, edit and save your code
 The 'Console' space allows you to run your code and see R's output
 The 'Environment' space is where R studio stores data files, variables and functions
 The last space holds several tabs as follows:
 'Files' to see files in your current working directory
 'Plots' any plot code you run will appear here
 'Packages' shows those packages that you have installed and those loaded for your sesion
 'Help' when you need some guidance from specific packages and functions
 'Viewer' don't know, never used it...

## Working with R
```{r, eval = FALSE}
# Operator '#'
# anything after a '#' will not be read as an operating code by R

# Some algebra in R####
100+1
2 * 6
10 /      2
log(10)

# Help
?
?log
help
example
citation

# comparing numbers with logical symbols####
1 < 2
0 > 1
1 != 2
1 <= 2
1 >= 1

# store information using/making variables####
x <- 5
x
cos(x)
y <- 10*x
y

#challenge problem 1####
#what will the following code return?
x <- 1
x <- x + 1 
x

# memory management####
ls()
x
rm(x)
x

# Workspace functions 
ls, exists, rm
getwd, setwd
install.packages, library

# Always start by cleaning your working environment, 
rm(list=ls())

# Install and load packages
install.packages("rmarkdown")
library(rmarkdown)

install.packages("ggplot2") #great plotting package
library(ggplot2)

# Some packages that will help us analyze and plot our data
install.packages("scales")
install.packages("MuMIn")
library("scales")
library("MuMIn")

# Set your working directory
getwd()
setwd(dir = "/Users/Dudisimo/Documents/PhD/TAing/Rworkshop_MBE2017")

# Loading some data
barnacles <-read.csv("data/BarnacleData.csv", header=TRUE, sep=",", strip.white = TRUE) #remove the 'data/'

# Figuring out data type
str(barnacles)
head(barnacles)
names(barnacles)
summary(barnacles)

```

```{r, eval = FALSE}
# The first functions to learn
str

# Important operators and assignment
%in%
$, [, [[, head, tail, subset

# Comparison 
!=, ==, >, >=, <, <=
is.na

# Basic math
*, +, -, /, ^
abs
ceiling, floor, round
exp, log, log10, log2, sqrt

max, min, prod, sum
range
mean, median, cor, sd, var

# Logical & sets 
&, |, !, xor
all, any
which

# Vectors and matrices
c, matrix
length, dim, ncol, nrow
cbind, rbind
names, colnames, rownames

# Making vectors 
c
rep, rep_len
seq, seq_len, seq_along
rev
sample
(is/as).(character/numeric/logical/...)

# Lists & data.frames 
list
data.frame, as.data.frame
expand.grid

# Control flow 
if
for, while
ifelse

# Apply & friends
lapply, sapply
```


