---
title: "Intro to R for MBE2017"
author: "Luis Malpica Cruz"
date: "May 2017"
output: html_document
---

## The basics
 This is your working environment in R Studio, here you write, edit and save your code
 The 'Console' space allows you to run your code and see R's output
 The 'Environment' space is where R studio stores data files, variables and functions
 The last space holds several tabs as follows:
 'Files' to see files in your current working directory
 'Plots' any plot code you run will appear here
 'Packages' shows those packages that you have installed and those loaded for your sesion
 'Help' when you need some guidance from specific packages and functions
 'Viewer' don't know, never used it...

## Working with R
```{r, eval = FALSE}
# Operator '#'
# anything after a '#' will not be read as an operating code by R

# Some algebra in R####
100+1
2 * 6
10 /      2
log(10)

# Help
?
help
# Challenge problem, find what do the next commands do:
?log
help(example)
help(citation)

# Store information using/making variables####
x <- 5
x
cos(x)
y <- 10*x
y

# Challenge problem
# What will the following code return?
x <- 1
x <- x + 1 
x

# Reading and manipulating data####
cats <- read.table(file = "data/cats.txt", header=TRUE, sep=",")
cats

# Manipulating data
cats$weight[1]
cats$weight[2]
cats$weight[1] + cats$weight[2]

# Challenge problem:
#what will the following print out and why?
cats$weight[1] + cats$coat[2]

# Investigating data types####
typeof(cats$weight[1])
is(cats$weight[1])
typeof(cats$coat[2])
is(cats$coat[2])
str(cats)

# 5 main types of data in R####
typeof(3.14)
typeof(1L)
typeof(1)
typeof(TRUE)
typeof(FALSE)
typeof('SWC')
typeof(1+1i)

#Why to know all these???

```

## Setting up an R session
```{r}
# Start by cleaning your working environment####
rm(list=ls())

# Install and load packages####
install.packages("ggplot2") #great plotting package
library(ggplot2)

# Set your working directory####
getwd()
setwd(dir = "/Users/Dudisimo/Documents/PhD/TAing/Rworkshop_MBE2017")

# Loading data into R (csv file) ####
gappy <- read.table(file = "data/gapminder-FiveYearData.csv", header=TRUE, sep=",", strip.white = TRUE)
str(gappy)
typeof(gappy)
typeof(gappy$year)

# Plotting with ggplot####
ggplot(data = gappy, aes(x = lifeExp, y = gdpPercap)) + geom_point()
ggplot(data = gappy, aes(x = gdpPercap, y = lifeExp)) + geom_point()
ggplot(data = gappy, aes(x = year, y = gdpPercap, color =continent)) + geom_point()

ggplot(data = gappy, aes(x = year, y = gdpPercap, by=country, color =continent)) + geom_point() + geom_line()

ggplot(data = gappy, aes(x = year, y = gdpPercap, by=country)) + geom_line(aes(color=continent))

ggplot(data = subset(gappy, continent=="Africa"), aes(x = year, y = gdpPercap, by=country)) +
  geom_line(aes(color=country))

ggplot(data = subset(gappy, continent=="Africa" & gdpPercap>3000), aes(x = year, y = gdpPercap, by=country)) + 
  geom_line(aes(color=country))

```
